Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-02-08 infercnv/reports/gains_and_losses.Rmd
write output to infercnv output dir to take advantage of cacheing

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  infercnv_dir: NULL
  rerun: true
---

@@

# chunk options
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE)

@@

# load object
obj <- 
  readRDS('/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/all_cells_reference.rds')

# get umap embeddings
umap <-
  obj$umap$embedding %>%
  tibble::as_tibble() %>%
  dplyr::bind_cols(
    # add metadata
    obj$meta_data %>%
      tibble::as_tibble() %>%
      dplyr::mutate(id = sample)) 
```

Next, we add the CNV metadata from the inferCNV run. We must wrangle it from wide- to long-format.

```{r infercnv_output}
# get cnv metadata
cnvs <- 
  list.files(
    infercnv_dir,
    pattern = 'map_metadata_from_infercnv.txt',
    recursive = T) %>%
  purrr::set_names(., dirname(.)) %>%
  purrr::map(function(file) {
    file.path(infercnv_dir, file) %>% 
      read.table(sep = '\t') %>% 
      tibble::as_tibble(rownames = 'cell')
    }) %>%
  dplyr::bind_rows(.id = 'id') %>%
  # convert from wide to long format
  tidyr::pivot_longer(
    cols = tidyselect::starts_with(c('has_', 'proportion_', 'top_'))
  ) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    chr = name %>%
      gsub('.*_', '', .) %>%
      {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
    event = sub(paste0(chr, '_'), '', name),
    measure = sub(paste0(chr, '_', event), '', name)) %>%
  dplyr::ungroup()

# combine umap and cnvs
umap <-
  umap %>%
  # add cnv metadata
  dplyr::inner_join(cnvs, by = c('id', 'cell')) 
```

@@

We will define whole chromosome events as those affecting >90% of genes in a given chromosome.

```{r call_ploidy}

@@

```{r}
# function: plot gains and losses
plot_gains_and_losses <- 
  function(umap, annot) {
    umap %>%
      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
      dplyr::filter(measure == 'has', event != 'cnv') %>%
      dplyr::group_by(id, chr, event, annotation) %>%
      dplyr::summarise(n_cells = dplyr::n(),
                       n_events = sum(value),
                       prop_of_cells = n_events / n_cells) %>%
      ggplot(aes(x = id, y = chr, alpha = prop_of_cells, fill = event)) +
      geom_tile() +
      theme_classic() +
      theme(panel.spacing = unit(0, "lines")) +
      scale_fill_manual(values = c('loss' = scales::muted('blue'),
                                   'dupli' = scales::muted('red'))) +
      scale_alpha(range = c(0, 1)) +
      facet_grid(event ~ annotation)
  }

plot_gains_and_losses(umap, 'cell_type')
plot_gains_and_losses(umap, 'cluster_name')

# plot chr7 gain on umap
umap %>%
  dplyr::filter(measure == 'has', event == 'dupli', chr == 'chr7') %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type, alpha = has_dupli_chr7)) +
  geom_point(size = 0.1) +
  theme_classic() +
  coord_fixed()
```

>>>>

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  infercnv_dir: NULL
  cache_dir: NULL
  rerun: true
---

@@

# chunk options
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = params$cache_dir)

@@

# load object
obj <- 
  readRDS('/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/all_cells_reference.rds')
```

Next, we add the CNV metadata from the inferCNV run to the UMAP embeddings. We must wrangle it from wide- to long-format.

```{r infercnv_output}
cnvs <- xfun::cache_rds({

  # get umap embeddings
  umap <- 
    obj$umap$embedding %>%
    tibble::as_tibble() %>%
    dplyr::bind_cols(
      # add metadata
      obj$meta_data %>%
        tibble::as_tibble() %>%
        dplyr::mutate(id = sample))

  # get cnv metadata
  cnvs <- 
    list.files(
      infercnv_dir,
      pattern = 'map_metadata_from_infercnv.txt',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(infercnv_dir, file) %>% 
        read.table(sep = '\t') %>% 
        tibble::as_tibble(rownames = 'cell')
      }) %>%
    dplyr::bind_rows(.id = 'id') %>%
    # convert from wide to long format
    tidyr::pivot_longer(
      cols = tidyselect::starts_with(c('has_', 'proportion_', 'top_'))
    )

  # get chr / event / name
  names <-
    cnvs %>%
    dplyr::distinct(name) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      chr = gsub('.*_', '', name),
      event = gsub(paste0('_', chr), '', name) %>% sub('.*_', '', .),
      measure = sub(paste0('_', event, '_', chr), '', name),
      chr = dplyr::case_when(
        measure == 'top' ~ paste0('chr', chr), TRUE ~ chr) %>% 
        {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())}) %>%
    dplyr::ungroup()

  # combine umap, names, and cnvs
  cnvs <-
    umap %>%
    # add cnv metadata
    dplyr::inner_join(cnvs, by = c('id', 'cell')) %>%
    # add names
    dplyr::left_join(names, by = 'name')

  # return
  cnvs

}, file = 'cnvs.rds')
```

@@

```{r}
# function: plot gains and losses
plot_gains_and_losses <- 
  function(cnvs, annot) {
    cnvs %>%
      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
      dplyr::filter(measure == 'has', event != 'cnv') %>%
      dplyr::group_by(id, chr, event, annotation) %>%
      dplyr::summarise(n_cells = dplyr::n(),
                       n_events = sum(value),
                       prop_of_cells = n_events / n_cells) %>%
      ggplot(aes(x = id, y = chr, alpha = prop_of_cells, fill = event)) +
      geom_tile() +
      theme_classic() +
      theme(panel.spacing = unit(0, "lines")) +
      scale_fill_manual(values = c('loss' = scales::muted('blue'),
                                   'dupli' = scales::muted('red'))) +
      scale_alpha(range = c(0, 1)) +
      facet_grid(event ~ annotation)
  }

plot_gains_and_losses(cnvs, 'cell_type')
plot_gains_and_losses(cnvs, 'cluster_name')

# plot celltype
cnvs %>%
  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
  geom_point(size = 0.1) +
  theme_classic() +
  coord_fixed()

# plot subclusters on umap
cnvs %>%
  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = subcluster)) +
  geom_point(size = 0.1) +
  theme_classic() +
  coord_fixed()

# plot chr7 gain on umap
cnvs %>%
  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) +
  geom_point(size = 0.1) +
  theme_classic() +
  scale_colour_viridis() +
  coord_fixed()
```

#### chr7 gain

We are most interested in the chr7 gain events in fibroblasts.

We will define cells with chr7 gain as those affecting >90% of genes.

```{r chr7_prop_dist}
p_dat <-
  cnvs %>%
  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
  dplyr::group_by(id, cluster_name, cell_type) %>%
  dplyr::mutate(median_proportion = median(value)) %>%
  dplyr::ungroup()

p_dat %>%
  ggplot(aes(x = value)) + 
  geom_density() +
  geom_vline(data = p_dat, aes(xintercept = median_proportion)) +
  ggh4x::facet_nested(~ subcluster + cell_type)
```

##### Pseudobulk

We pseudobulk chr7+ cells and perform differential expression analysis.

```{r pseudobulk}

@@
